---
title: 'CSCI E-63C: Final Exam'
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preface

For the final exam/project we will develop classification models using several approaches and compare their performance on a new dataset -- so-called "Census Income" from UCI ML.  It is available at UCI ML web site, but so that we are not at the mercy of UCI ML availability, there is also a local copy of it in our website in Canvas as a zip-archive of all associated files.  Among other things, the description for this dataset also presents performance (prediction accuracy) observed by the dataset providers using variety of modeling techniques -- this supplies a context for the errors of the models we will develop here.

Please note that the original data has been split up into training and test subsets, but there doesn't seem to be anything particular about that split, so we might want to pool those two datasets together and split them into training and test as necessary ourselves. As you do that, please check that the attribute levels are consistent between those two files.  For instance, the categorized income levels are indicated using slightly different notation in their training and test data.   By now it should be quite straightforward for you to correct that when you pool them together.

Also, please note that there is non-negligible number of rows with missing values that for most analyses cannot be included without modification in the computation.  Please decide how you want to handle them and proceed accordingly.  The simplest and perfectly acceptable approach would be to exclude those observations from the rest of the analyses, but if you have time and inclination to investigate the impact of imputing them by various means, you are welcome to try.

Attribute called "final weight" in the dataset description represents demographic weighting of these observations.  Please disregard it for the purposes of this assignment.

Additionally, several attributes in this dataset are categorical variables with more than two levels (e.g. native country, occupation, etc.).  Please make sure to translate them into corresponding sets of dummy indicator variables for the methods that require such conversion (e.g. PCA) -- R function `model.matrix` can be convenient for this, instead of generating those 0/1 indicators for each level of the factor manually (which is still perfectly fine).  Some of those multi-level factors contain very sparsely populated categories -- e.g. occupation "Armed-Forces" or work class "Never-worked" -- it is your call whether you want to keep those observations in the data or exclude also on the basis that there is not enough data to adequately capture the impact of those categories. Feel free to experiment away!

Among the multi-level categorical attributes, native country attribute has the largest number of levels -- several folds higher than any other attribute in this dataset -- some of which have relatively few observations.  This associated increase in dimensionality of the data may not be accompanied by a corresponding gain of resolution -- e.g. would we expect this data to support the *difference* in income between descendants from Peru and Nicaragua, for example, or from Cambodia and Laos?  Please feel free to evaluate the impact of inclusion and/or omission of this attribute in/from the model and/or discretizing it differently (e.g. US/non-US, etc.).

Lastly, the size of this dataset can make some of the modeling techniques run slower than what we were typically encountering in this class.  You may find it helpful to do some of the exploration and model tuning on multiple random samples of smaller size as you decide on useful ranges of parameters/modeling choices, and then only perform a final run of fully debugged and working code on the full dataset.

```{r child = 'problem1.Rmd'}
```

```{r child = 'problem2.Rmd'}
```

```{r child = 'problem3.Rmd'}
```

```{r child = 'problem4.Rmd'}
```

```{r child = 'problem5.Rmd'}
```

```{r child = 'extra1.Rmd'}
```

```{r child = 'extra2.Rmd'}
```


# Problem 3: random forest (25 points)

Develop random forest model of the categorized income. Present variable importance plots and comment on relative importance of different attributes in the model.  Did attributes showing up as more important in random forest model also appear as significantly associated with the outcome by logistic regression?  Test model performance on multiple splits of data into training and test subsets, compare test and out-of-bag error estimates, summarize model performance in terms of accuracy/error, sensitivity/specificity and compare to the performance of other methods reported in the dataset description.
```{r}
#Trying out with 1000 trees and varying levels of mtry
set.seed(1234)
rfRes <- randomForest(class~., ntree=1000, data = train,mtry=c(1,2,5,10,20,50,100,200))

table(train$class,predict(rfRes))
assess(train$class==">50K",predict(rfRes)==">50K")

print(rfRes)
plot(rfRes)

varImpPlot(rfRes)

tune.rf <- tuneRF(train[,-12],train[,12], stepFactor=0.5)
plot(tune.rf)

#Using mtry=3 and ntree=200
set.seed(345)
rfTmp <-NULL
for ( iSim in 1:30) {
    bTrain <- sample(c(FALSE,TRUE,TRUE),nrow(train),replace=TRUE)
    rfTrain <- randomForest(class~.,data=train[bTrain,],ntree=200,mtry=3)
    rfTestPred <- predict(rfTrain, newdata=train[!bTrain,]) 
    tmpVals <- assess(train[!bTrain,]$class==">50K",rfTestPred==">50K")
    rfTmp <- rbind(rfTmp,data.frame(iSim,tmpVals,mean.oob= mean(rfTrain$err.rate[,1])*100))
}


melt(rfTmp,id.vars="iSim") %>% ggplot(aes(x=variable,y=value,colour=variable)) + geom_boxplot() + ggtitle("Random Forest")


perf <- rbind(perf, data.frame(algo="Random Forest",err=mean(rfTmp$error)))
kable(perf)

ggplot(perf,aes(x=algo,y=err)) + geom_point() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +labs(x="Algorithm",y="Error")

```

# Problem 4: SVM (25 points)

Develop SVM model of this data choosing parameters (e.g. choice of kernel, cost, etc.) that appear to yield better performance.  Test model performance on multiple splits of data into training and test subsets, summarize model performance in terms of accuracy/error, sensitivity/specificity and compare to the performance of other methods reported in the dataset description.
```{r}

#Using a sample of the dataset for parameter tuning as it takes too long. Its an old laptop :( 
set.seed(34343)
ttrain <- train[sample(nrow(train), 3000), ]


#First trying with a linear kernel and diffent costs
summary(tune(svm,class~.,data=ttrain,kernel="linear",ranges=list(cost=c(0.1,1,2,5,10,20,50))))

#Trying the radial kernel with different values for cost and gamma parameters
tune(svm,class~.,data=ttrain,kernel="radial",ranges=list(cost=c(.5,1,2,5,10),gamma=c(0.05,0.1,1.5)))
  

#We get about an error reduction of about 1 percent with the radial kernel (cost=1, gamma=.1) vs the linear kernel
  

#Running bootstrap with our best observed model
set.seed(345)
svTmp <-NULL
for ( iSim in 1:30) {
    bTrain <- sample(c(FALSE,TRUE,TRUE),nrow(train),replace=TRUE)
    svTrain <- svm(class~.,data=train[bTrain,],kernel="radial",cost=1,gamma=.1)
    svTestPred <- predict(svTrain, newdata=train[!bTrain,]) 
    tmpVals <- assess(train[!bTrain,]$class==">50K",svTestPred==">50K")
    svTmp <- rbind(svTmp,data.frame(iSim,tmpVals))
}

melt(svTmp,id.vars="iSim") %>% ggplot(aes(x=variable,y=value,colour=variable)) + geom_boxplot() + ggtitle("SVM")


perf <- rbind(perf, data.frame(algo="SVM",err=mean(svTmp$error)))
kable(perf)

ggplot(perf,aes(x=algo,y=err)) + geom_point() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +labs(x="Algorithm",y="Error")

```

# Problem 5: compare logistic regression, random forest and SVM model performance (5 points)

Compare performance of the models developed above (logistic regression, random forest, SVM) in terms of their accuracy, error and sensitivity/specificity.  Comment on differences and similarities between them.


```{r}

tot <- rbind(lrTmp,rfTmp[,-1],svTmp)
tot$lbl <- c(rep("log regression",30),rep("random forest",30),rep("SVM",30))

tot %>% melt(id.vars="lbl") %>% ggplot(aes(x=lbl,y=value,group=variable)) + geom_boxplot() + facet_wrap (~variable)


#Going to perform on the holdout test set defined in Question 1

rocplot=function(pred, truth,...){
  predob = prediction (pred, truth)
  perf = performance (predob , "tpr", "fpr")
  plot(perf,...)
}




```

# Extra 10 points: KNN model

Develop KNN model for this data, evaluate its performance for different values of $k$ on different splits of the data into training and test and compare it to the performance of other methods reported in the dataset description.  Notice that this dataset includes many categorical variables as well as continuous attributes measured on different scales, so that the distance has to be defined to be meaningful (probably avoiding subtraction of the numerical values of multi-level factors directly or adding differences between untransformed age and capital gain/loss attributes).
```{r}

 
dfTmp <- NULL
for ( iSim in 1:30 ) {
  trainIdx <- sample(nrow(dbaDat),nrow(dbaDat),replace=TRUE)
  knnTuneRes <- tune.knn(dbaDat[trainIdx,-ncol(dbaDat)],dbaDat[trainIdx,ncol(dbaDat)],k=1:10)
  knnTestRes <- knn(dbaDat[trainIdx,-ncol(dbaDat)],dbaDat[-trainIdx,-ncol(dbaDat)],dbaDat[trainIdx,ncol(dbaDat)],k=knnTuneRes$best.parameters[,"k"])
  tblTmp <- table(dbaDat[-trainIdx,"auth"],knnTestRes)
  #print(tblTmp)
  dfTmp <- rbind(dfTmp,data.frame(attr=c("k","err0","err1","errTot"),value=c(knnTuneRes$best.parameters[,"k"],tblTmp[1,2]/sum(tblTmp[1,]),tblTmp[2,1]/sum(tblTmp[2,]),1-sum(diag(tblTmp))/sum(tblTmp))))
}

```

# Extra 15 points: variable importance in SVM

SVM does not appear to provide readily available tools for judging relative importance of different attributes in the model.  Please evaluate here an approach similar to that employed by random forest where importance of any given attribute is measured by the decrease in model performance upon randomization of the values for this attribute.
```{r}

library(caret)




```

