---
title: "problem1"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(dplyr)
library(reshape2)
library(ggplot2)
library(countrycode)
library(randomForest)
library(stringr)
library(e1071)
```
# Problem 1: univariate and unsupervised analysis (20 points)

Download and read "Census Income" data into R and prepare graphical and numerical summaries of it: e.g. histograms of continuous attributes, contingency tables of categorical variables, scatterplots of continuous attributes with some of the categorical variables indicated by color/symbol shape, etc.  Perform principal components analysis of this data (do you need to scale it prior to that? how would you represent multilevel categorical attributes to be used as inputs for PCA?) and plot observations in the space of the first few principal components with subjects' gender and/or categorized income indicated by color/shape of the symbol.  Perform univariate assessment of associations between outcome we will be modeling and each of the attributes (e.g. t-test or logistic regression for continuous attributes, contingency tables/Fisher exact test/$\chi^2$ test for categorical attributes).  Summarize your observations from these assessments: does it appear that there is association between outcome and predictors? Which predictors seem to be more/less relevant?

```{r}

adultDat <- read.csv("adult.data", header = FALSE)

#Skipping first row as it contains a strange string
adultTest <- read.csv("adult.test", skip=1, header=FALSE)

#Take a peek
head(adultDat)
head(adultTest)

#Assign names, taken from the corresponding txt files
adultNames <- unique(str_match(readLines("old.adult.names.txt"), "^[a-zA-Z-]+"))
adultNames <- adultNames[!is.na(adultNames)]

#Replace dashes with dots for easier reference
adultNames <-gsub('-','.',adultNames)

#Assign names 
colnames(adultDat) <- adultNames
head(adultDat)

colnames(adultTest) <- adultNames
head(adultTest)

#Check that numerics are read in correctly and characters read as factors
summary(adultDat)
summary(adultTest)

#From here we can see that the numerics values are clean (no NAs). 
#Will transpose and merge the unique values of each character variable
#to see what may need cleaning before we concatenate the datasets
sapply(adultDat,class)

#Store bool of Character and Numeric variables
charVect <- sapply(adultDat, class) %in% c('character', 'factor')
numVect <- sapply(adultDat, class) %in% c('integer')

#Distinct categorical data in each data frame
adultLong <- melt(as.matrix(adultDat[,charVect])) %>% group_by(Var2,value) %>% summarise(n.adult=n_distinct(value)) 
testLong <- melt(as.matrix(adultTest[,charVect])) %>% group_by(Var2,value) %>% summarise(n.test=n_distinct(value)) 

#Values that will need to be investigated
merge(adultLong,testLong, by=c("Var2","value"),all=TRUE) %>% filter(is.na((n.adult+n.test)))

#Holland-Netherlands is a valid country name, so will just compress the . from the class variable in the test dataset.
adultTest$class <- gsub('[.]','',adultTest$class)

#We can also see that we have some missing values in the following variables that we will have to look into imputing
adultLong[adultLong$value==' ?',1]

#Join our cleaned datasets
incomeDat <- rbind(adultDat,adultTest)

#Checking that the class of variables is preserved
summary(incomeDat)

#Class looks to fixed now
missWrkClss <- incomeDat[incomeDat$workclass==' ?',]
missOcupatn <- incomeDat[incomeDat$occupation==' ?',]
missNatCntr <- incomeDat[incomeDat$native.country==' ?',]

#For the time being, going to replace vars with ? with NA
incomeDat[,charVect] <- apply(incomeDat[,charVect],2,function(x) gsub(' [?]',NA,x))

#Remove the leading spaces from the character variables
incomeDat[,charVect] <- apply(incomeDat[,charVect],2,function(x) gsub('^ ','',x))

#Converting back to factor
incomeDat <- as.data.frame(unclass(incomeDat))

#Remove all records with NA
nIncomeDat <- na.omit(incomeDat)

#Lets see how much data we lose by removing NAs
(nrow(incomeDat)-nrow(nIncomeDat))/nrow(incomeDat)*100

#We lose about 7.5% of the data, however it seems more hassle than its worth to impute values for this amount of data, so will continue with these. Especially since we cannot guarantee whether the missings are random or if there is an underlying pattern.

#Dataset is too big for testing, taking a random sample of 2k 
sampIncome <- data.frame(nIncomeDat[sample(1:nrow(nIncomeDat), 2000,
  	replace=FALSE),])

#Categorical data
#Contingency table
nIncomeDat[,charVect] %>% melt(id="class") %>% group_by(variable,value,class) %>% summarise(n=n()) %>% dcast(variable+value~class,value.var="n") %>% kable()

#We have a lot of granularity in the categorical data which we could do without

#Looking at the education level and education num, it looks like we dont need both education and education.num. Furthermore its prob better to collapse the levels also and turn it into dummy variables if possible
table(sampIncome$education,sampIncome$education.num)

table(sampIncome$marital.status,sampIncome$relationship)
#Bar char colored by outcome
sampIncome[,charVect] %>% melt(id.vars="class") %>% 
  ggplot(aes(x=value,fill=class)) + geom_bar() + facet_wrap(~variable, scales = "free") 


#Performing Chi-sq test on the whole dataset to see which variables are independent of income class
lapply(nIncomeDat[,charVect], function(x) {
chisq.test(x, nIncomeDat$class)
})  
#It seems that all variables are related to the outcome variable with a very strong significance



#Continuous variables
#Creates color pallette based on unique column values
#taken from ISLR
Cols=function(vec){
  cols=rainbow(length(unique(vec)))
  return(cols[as.numeric(as.factor(vec))]) 
}

#Scatterplot matrix, colored by income class
pairs(~.,data=sampIncome[,numVect], 
        col=Cols(sampIncome$class))

#Here age seems to be correlated with weight (fnlwgt) as is expected


#Histogram plot
as.matrix(sampIncome[,numVect]) %>% melt() %>% mutate(value=as.numeric(value)) %>% ggplot(aes(x=value)) + geom_histogram() + facet_wrap(~Var2, scales = "free")


lapply(nIncomeDat[,numVect], function(x) t.test(x ~ nIncomeDat$class, var.equal = TRUE))
#fnlWgt is the only variable that has no relationship with the outcome variable. We may consider dropping this


nvClass <- c(numVect[1:14],TRUE)

sampIncome[,nvClass] %>% melt(id.vars="class") %>% ggplot(aes(x=variable, y=value,group=class,fill=class)) + geom_boxplot() + facet_wrap(~variable, scales = "free")

#drop fnlwgt
#log transform age
#drop education.num
#pre-college (1:10), 

plot(nIncomeDat$education.num,col=Cols(nIncomeDat$class))
#Although education matters, it doesnt seem to be a large distinction prior to high school


cIncomeDat <- nIncomeDat
cIncomeDat$education <- as.character(cIncomeDat$education)
cIncomeDat$education[cIncomeDat$education.num <=10] <- "Pre-College"
cIncomeDat$education<- as.factor(cIncomeDat$education)

#Drop education num
cIncomeDat$education.num <- NULL

#Drop final weight
cIncomeDat$fnlwgt <- NULL

cIncomeDat$age <- log(cIncomeDat$age)

#Recoding countries to regions
cIncomeDat$region <- countrycode(cIncomeDat$native.country,"country.name","region")

#Some issues converting columbia, england, hong, scotland, south and yugoslavia
#need to fix these and try again

cIncomeDat$native.country <- as.character(cIncomeDat$native.country)
cIncomeDat$native.country[cIncomeDat$native.country=="Columbia"] <- "Colombia"
cIncomeDat$native.country[cIncomeDat$native.country=="Hong"] <- "Hong Kong"
cIncomeDat$native.country[cIncomeDat$native.country %in% c("England","Scotland")] <- "UK"

#For the sake of argument, just picking one
cIncomeDat$native.country[cIncomeDat$native.country %in% c("Yugoslavia")] <- "Bosnia"

cIncomeDat <- cIncomeDat[cIncomeDat$native.country!="South",]
cIncomeDat$region <-( countrycode(cIncomeDat$native.country,"country.name","region"))
cIncomeDat$region <- as.factor(cIncomeDat$region)
cIncomeDat$native.country <- NULL

#Creating our data matrix that will be used for the rest of the analysis
nrow(cIncomeDat)


dm <- model.matrix(~.+0, data=cIncomeDat)

#We have one variable (workclass never worked) with no values, we will remove this
dm <- dm[,-4]

summary(dm)

pr <- prcomp(dm,scale=TRUE)
par(mfrow=c(1,1))
biplot(pr,cex=.5)
#From the biplot, it looks like Sex, and Region contribute heavily to the second principal component

plot(pr)
#We can see that the first two principal components to a good job of explaining a lot of the variance. However, we see that the data may not be so easily separable after that

plot(pr$x[,1:2], col=Cols(cIncomeDat$class))
#The class seems to be rather well defined by the second principal component

plot(pr$x[,1:2], col=Cols(cIncomeDat$sex))
#Gender is rather well defined by the second principal component but not nearly as well separated as class

#Top 5 contributers to PC1 and PC2 respectively
sort(abs(pr$rotation[,1]),decreasing=T)[1:5]
sort(abs(pr$rotation[,2]),decreasing=T)[1:5]

set.seed(5124)

#holding out 20% for the test set
gTrain <- sample(c(FALSE,TRUE,TRUE,TRUE,TRUE),nrow(cIncomeDat),replace=TRUE)

train <- cIncomeDat[gTrain,]
test <- cIncomeDat[!gTrain,]

```

```{r include=FALSE}

#Writing out for use of other problems
write.table(train,file="train.csv")
write.table(test,file="test.csv")
```

